Automatic Identification and Annotation of Parallel Tasks in Structured Programs

### 2017-11-17 docs/ACM/paper.tex
Adding examples to section 2.
Discussing the challenges.

%% Packages that we are using in this version of the work:
\usepackage{hyperref}

@@

@@

This paper describes \Taskminer, a compiler that mines task parallelism
from programs.
To fulfill this goal \Taskminer{} solves many challenges.
First, it finds program regions, e.g., loops or functions, that can be
effectively mapped onto tasks (Section~\ref{sub:region}).
Second, it determines symbolic bounds to all the memory blocks accessed within
those regions (Section~\ref{sub:symb}).
Third, it maps such information back into source code, producing annotations
that programmers can read (Section~\ref{sub:ir}).
Fourth, it extracts parameters from the code to estimate the profitability of
tasks, given the cost to create them.
These parameters feed conditional checks, which, at runtime, enable or disable
the creation of tasks (Section~\ref{sub:profit}).
Finally, \Taskminer{} determines which program variables need to be replicated at
each new task, or shared among them (Section~\ref{sub:variance}).
This paper describes the design and the implementation of these technologies, and
demonstrate their benefit.

@@

@@

\subsection{Mapping Program Regions to Tasks}
\label{sub:region}

@@

\subsection{Estimating the Profitability of Tasks}
\label{sub:ir}

\subsection{Separating Private from Shared Variables}
\label{sub:profit}

\subsection{Mapping IR onto Source Code}
\label{sub:variance}

>>>>

%% Packages that we are using in this version of the work:
\usepackage{hyperref}
\usepackage{amsmath}

@@

\newtheorem{Challenge}{Challenge}[section]

@@

This paper describes \Taskminer, a compiler that mines task parallelism
from programs.
To fulfill this goal \Taskminer{} solves many challenges.
First, it finds program regions, e.g., loops or functions, that can be
effectively mapped onto tasks (Section~\ref{sub:identification}).
Second, it determines symbolic bounds to all the memory blocks accessed within
those regions (Section~\ref{sub:symb}).
Third, it maps such information back into source code, producing annotations
that programmers can read (Section~\ref{sub:ir}).
Fourth, it extracts parameters from the code to estimate the profitability of
tasks, given the cost to create them.
These parameters feed conditional checks, which, at runtime, enable or disable
the creation of tasks (Section~\ref{sub:profit}).
Finally, \Taskminer{} determines which program variables need to be replicated at
each new task, or shared among them (Section~\ref{sub:variance}).
This paper describes the design and the implementation of these technologies, and
demonstrate their benefit.

@@

The automatic insertion of effective OpenMP annotations into C/C++ programs
required us to deal with a number of challenges.
If left unsolved, these challenges would restrict our interventions in programs
to very trivial annotations, which would hardly be of any use.
In the rest of this section, we discuss the most important challenges, using
this discussion as a motivation to the developments that follow in
Section~\ref{sec:sol}.

\begin{Challenge}
\label{ch:Regions}
Identify the memory region covered by a task.
\end{Challenge}

Figure~\ref{fig:ex_Regions} illustrates Challenge~\ref{ch:Regions}, and shows
how we solve it~\footnote{Figures~\ref{fig:ex_Regions}-\ref{fig:ex_cutoff} shows
programs that we have annotated without any human intervention.
Statements in black font are part of the original program.
Annotations appear in grey.}.
That program receives an $\mathsf{M}\times\mathsf{N}$ matrix \textsf{V}, in
linearized format, and produces a vector \textsf{U}, so that \textsf{U[i]}
contains the sum of all the elements in line \textsf{i} of matrix \textsf{V}.
For reasons that will be made more clear in Section~\ref{sub:identification},
our static analysis determines that each iteration of the outermost loops
could be made into a task.
Thus, tasks comprise the innermost loop, and traverse the memory region
between addresses $\textsf{\&V + i * N}$ and $\textsf{\&V + i * N + M}$.
The identification of such ranges involves the use of a symbolic algebra,
which we have borrowed from the compiler-related literature, as we
explain in Section~\ref{sub:symb}.
Figure~\ref{fig:ex_Regions} also lets us introduce the second challenge that we
tackle in this paper:

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.95\columnwidth]{images/ex_Regions}
\caption{Write caption here.}
\label{fig:ex_Regions}
\end{center}
\end{figure}

\begin{Challenge}
\label{ch:cost}
Estimate the profitability of tasks at runtime.
\end{Challenge}

The creation of tasks involves a non-negligible runtime cost, which includes
the allocation of resources, and the scheduling of the task proper.
Ideally, this cost should be paid only when the task performs an amount of
work that is sufficiently large to pay for its management cost.
Being an interesting program property, on Rice's sense~\cite{Rice53}, the amount
of work performed by a task cannot be discovered statically.
However, we can try to approximate this quantity, using, to this end, program
symbols, which are replaced with actual values at runtime.
For instance, in Figure~\ref{fig:ex_Regions}, we know that the body of the
innermost loop is formed by five instructions.
Thus, we approximate the amount of work performed by a task with the
expression \textsf{5 * M}.
We use the runtime value of \textsf{M} to determine, during the execution of the
program, if we create a task, or not.
Such test is carried out by the guard at line 7 of the figure, which is part of
OpenMP's syntax.

\begin{Challenge}
\label{ch:cutoff}
Bound the creation of recursive tasks.
\end{Challenge}

To introduce Challenge~\ref{ch:cutoff}, we quote Duran {\em et al.}:
``{\em In task parallel languages, an important factor for achieving a good
performance is the use of a cut-off technique to reduce the number of tasks
created}"~\cite{Duran08}.
This observation is particularly true in the context of recursive, fine-grain,
tasks.
Figure~\ref{fig:ex_cutoff} provides such an example.
To place a limit on the number of tasks simultaneously in flight, we associate
the invocation of recursive functions annotated with task pragmas with a
counter -- \textsf{taskminer\_depth\_cutoff} in Figure~\ref{fig:cutoff}.
The guard in line 7 ensures that we never exceed \textsf{DEPTH\_CUTOFF}, a
predetermined threshold.
The code generation algorithms that we discuss in this paper are
parameterized by constants such as \textsf{DEPTH\_CUTOFF}, or
\textsf{WORK\_CUTOFF} in Figure~\ref{fig:ex_Regions}.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=0.95\columnwidth]{images/ex_cutoff}
\caption{Write caption here.}
\label{fig:ex_cutoff}
\end{center}
\end{figure}

\begin{Challenge}
\label{ch:privatize}
Identify private and shared variables.
\end{Challenge}

The two previous challenges are related to the performance of annotated
programs: if left unsolved, we shall have correct, albeit inefficient programs.
Challenges~\ref{ch:Regions}, and~\ref{ch:privatize}, on the other hand, are
related to correctness.
Challenge~\ref{ch:privatize} asks for the identification of variables that
must be replicated among threads.
This process of replication is called {\em privatization}.
As an example, variable \textsf{j} in Figure~\ref{fig:ex_privatize} must
be privatized.
In the absence of such action, that variable will be shared among all the
tasks created at line 5 of Figure~\ref{fig:ex_privatize}.
Because \textsf{j} is writte within these tasks, race conditions would
ensue.
In Section~\ref{sub:variance} we explain how we separate private from
shared variables.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\columnwidth]{images/ex_privatize}
\caption{Write caption here.}
\label{fig:ex_privatize}
\end{center}
\end{figure}

@@

\subsection{Mapping Program Regions to Tasks}
\label{sub:identification}

@@

\subsection{Estimating the Profitability of Tasks}
\label{sub:profit}

\subsection{Separating Private from Shared Variables}
\label{sub:variance}
